\section{Introduction}
  \label{sec:introduction}

  The physical notions of mass, electric charge, and inertia occupy a central role in modern physics.
  They enter as fundamental parameters in both classical and quantum theories, yet their conceptual
  origin remains largely unexplained.

  In the Standard Model, mass and charge are introduced as intrinsic properties of elementary fields,
  with numerical values fixed by empirical input.
  While the Higgs mechanism provides a dynamical account of mass generation for gauge bosons and fermions,
  it does not address why mass exists as a property, nor why inertial response accompanies it~\cite{Higgs1964}.
  Electric charge, similarly, is treated as a primitive coupling constant, constrained by gauge symmetry
  but not derived from a deeper structural principle~\cite{PeskinSchroeder}.

  From a foundational perspective, inertia presents an additional conceptual challenge.
  The resistance of a system to acceleration is usually taken as a defining feature of matter,
  yet its relation to gravitational mass and its possible emergence from more primitive structures
  have been the subject of long-standing debate~\cite{Mach1893,Einstein1916}.

  A parallel line of inquiry has emerged in approaches to quantum gravity and pre-geometric physics.
  In these frameworks, spacetime geometry, and sometimes even locality and causality, are not assumed
  as fundamental, but are reconstructed as effective descriptions from more primitive relational or
  algebraic structures~\cite{Rovelli2004,AmelinoCamelia2013}.
  Within such approaches, familiar physical quantities may acquire an emergent or regime-dependent status.

  Recent work has emphasized that effective physical descriptions often arise through projections
  from an underlying configuration space to observable spacetime variables.
  When such projections are non-injective, multiple underlying configurations correspond to the same
  effective observable state, leading to intrinsic information loss at the descriptive level~\cite{Beau2026b}.
  This structural feature has been shown to account for the breakdown of classical probabilistic
  factorization in quantum correlations, without invoking nonlocal dynamics~\cite{Bell1964}.

  In parallel, bounded-response mechanisms have been extensively studied in effective field theories.
  Born--Infeld electrodynamics provides a paradigmatic example in which divergences are regulated
  by intrinsic saturation of the field response~\cite{BornInfeld1934}.
  Related saturation phenomena appear in gravitational and cosmological contexts, where effective
  responses cease to scale linearly beyond certain thresholds.

  Motivated by these developments, we investigate whether mass, electric charge, and inertial response
  can be understood as manifestations of a single structural mechanism.
  Specifically, we explore a pre-geometric relational framework in which physical observables arise
  through a projection subject to intrinsic saturation bounds.

  Within this setting, mass is interpreted as an isotropic inhibition of relational relaxation,
  while electric charge corresponds to an oriented or asymmetric saturation of the same underlying
  relational flux.
  Inertia then emerges as a secondary effect, reflecting the finite resolvability of changes in
  relational configurations rather than a primitive dynamical property.

  The purpose of this work is not to modify the Standard Model or to introduce new fundamental fields.
  Instead, we aim to provide a structural reinterpretation of existing quantities, clarifying their
  mutual relations and their status as effective descriptors.
  By identifying a common origin for mass, charge, and inertia, the framework offers a unified
  conceptual basis for inertial and interaction properties within a projection-limited description.

  The paper is organized as follows.
  Section~\ref{sec:framework} introduces the relational and projective framework underlying the analysis.
  Section~\ref{sec:saturation} develops the classification of saturated effective responses and
  shows how mass, electric charge, and inertia arise as distinct symmetry realizations within this
  framework.
  Section~\ref{sec:consistency} examines the consistency of this interpretation with Standard Model phenomenology.
  We conclude in Section~\ref{sec:discussion} with a discussion of conceptual implications,
  limitations, and possible directions for further investigation.
